Bayesian Inference from Time Series: Information Loss and the Whittle Likelihood

### 2018-03-05 doc/ms.tex
More citations and intro writing

@@

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
\toprule
$N$         &       Likelihood      &  Mut. Inf. (nats) \\
\hline
20          &       Exact           & \\
20          &       Whittle         & $1.491 \pm 0.055$\\ 
50          &       Exact           & \\
50          &       Whittle         & $1.957 \pm 0.055$ \\
100         &       Exact           & $2.605 \pm 0.053$ \\
100         &       Whittle         & $2.052 \pm 0.055$ \\
200         &       Exact           &   \\
200         &       Whittle         &   \\
\bottomrule
\end{tabular}
\caption{The mutual information between the data and the log-period, for
different numbers of datapoints $N$ and for different likelihoods (exact vs.
Whittle).
This is equal to the prior expected reduction in the entropy
in going from the prior to the posterior.\label{tab:oscillator_results}}
\end{table}

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{}}
\toprule
$N$         &       $\Delta$ Mut. Inf. (nats) \\
\hline
50          & \\
100         &       $0.553 \pm 0.056$ \\
200         & \\
\bottomrule
\end{tabular}
\caption{The difference in mutual information between the exact and
Whittle likelihoods. This quantifies the size of the advantage of the exact
likelihood.
The error bars are smaller than what you might expect from
taking differences in Table~\ref{tab:oscillator_results}, due to
the common stream of datasets.\label{tab:oscillator_differences}}
\end{table}

\begin{thebibliography}{999}
\bibitem[Anderson et al.(1990)]{anderson}
Anderson, Edwin R., Thomas L. Duvall Jr, and Stuart M. Jefferies. ``Modeling of solar oscillation power spectra.'' The Astrophysical Journal 364 (1990): 699-705.

@@

\bibitem[Foreman-Mackey et al.(2017)]{celerite}
Foreman-Mackey, Daniel, Eric Agol, Sivaram Ambikasaran, and Ruth Angus. ``Fast and scalable Gaussian process modeling with applications to astronomical time series.'' The Astronomical Journal 154, no. 6 (2017): 220.
\end{thebibliography}

>>>>

In astronomy, time series analysis of variable sources has often been performed
in the frequency domain \citep[e.g.][]{lomb, hatzes, helio}. However,
a number of researchers have applied existing time series models from
statistics \citep{brockwell_davis} or custom-made models to infer parameters
directly from the time series data, without any use of the frequency domain
\citep[e.g.][]{gregory_loredo, gregory, kelly, brewer_stello, farr}.

Methods that analyse the time series data directly are often less convenient than frequency-domain fits, and can have a much greater computational cost,
despite recent computational advances which have improved the matter
considerably \citep[e.g.][]{hodlr, celerite}. Yet the question remains ---
if we opt for the convenient frequency-domain methods, how much relevant
information are we throwing away? That is the focus of this paper.

@@

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
\toprule
$N$         &       Likelihood      &  Mut. Inf. (nats) \\
\hline
20          &       Exact           & \\
20          &       Whittle         & $1.491 \pm 0.055$\\ 
50          &       Exact           & \\
50          &       Whittle         & $1.957 \pm 0.055$ \\
100         &       Exact           & $2.605 \pm 0.053$ \\
100         &       Whittle         & $2.052 \pm 0.055$ \\
200         &       Exact           &   \\
200         &       Whittle         &   \\
\bottomrule
\end{tabular}
\caption{The mutual information between the data and the log-period, for
different numbers of datapoints $N$ and for different likelihoods (exact vs.
Whittle).
This is equal to the prior expected reduction in the entropy
in going from the prior to the posterior.\label{tab:oscillation_results}}
\end{table}

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{}}
\toprule
$N$         &       $\Delta$ Mut. Inf. (nats) \\
\hline
50          & \\
100         &       $0.553 \pm 0.056$ \\
200         & \\
\bottomrule
\end{tabular}
\caption{The difference in mutual information between the exact and
Whittle likelihoods. This quantifies the size of the advantage of the exact
likelihood.
The error bars are smaller than what you might expect from
taking differences in Table~\ref{tab:oscillation_results}, due to
the common stream of datasets.\label{tab:oscillation_differences}}
\end{table}

\begin{thebibliography}{999}

\bibitem[Ambikasaran et al.(2016)]{hodlr}
Ambikasaran, Sivaram, Daniel Foreman-Mackey, Leslie Greengard, David W. Hogg, and Michael O’Neil. ``Fast direct methods for Gaussian processes.''
IEEE transactions on pattern analysis and machine intelligence 38, no. 2 (2016): 252-265.

\bibitem[Anderson et al.(1990)]{anderson}
Anderson, Edwin R., Thomas L. Duvall Jr, and Stuart M. Jefferies. ``Modeling of solar oscillation power spectra.'' The Astrophysical Journal 364 (1990): 699-705.

@@

\bibitem[Brockwell and Davis(2013)]{brockwell_davis}
Brockwell, P.J. and Davis, R.A., 2013. Time series: theory and methods. Springer Science \& Business Media.

\bibitem[Christensen-Dalsgaard(2002)]{helio}
Christensen-Dalsgaard, Jørgen. ``Helioseismology.'' Reviews of Modern Physics 74, no. 4 (2002): 1073.

\bibitem[Farr et al.(2018)]{farr}
Farr, Will M., Benjamin JS Pope, Guy R. Davies, Thomas SH North, Timothy R. White, Jim W. Barrett, Andrea Miglio et al. ``Aldebaran b's temperate past uncovered in planet search data.'' arXiv preprint arXiv:1802.09812 (2018).

\bibitem[Foreman-Mackey et al.(2017)]{celerite}
Foreman-Mackey, Daniel, Eric Agol, Sivaram Ambikasaran, and Ruth Angus. ``Fast and scalable Gaussian process modeling with applications to astronomical time series.'' The Astronomical Journal 154, no. 6 (2017): 220.

\bibitem[Gregory(2005)]{gregory}
Gregory, Philip C.
``A Bayesian analysis of extrasolar planet data for HD 208487.''
In AIP Conference Proceedings, vol. 803, no. 1, pp. 139-145. AIP, 2005.

\bibitem[Gregory and Loredo(1992)]{gregory_loredo}
Gregory, P. C., and Thomas J. Loredo. ``A new method for the detection of a periodic signal of unknown shape and period.'' The Astrophysical Journal 398 (1992): 146-168.

\bibitem[Hatzes et al.(1993)]{hatzes}
Hatzes, A.P. and Cochran, W.D., 1993. Long-period radial velocity variations in three K giants. The Astrophysical Journal, 413, pp.339-348.

\bibitem[Kelly(2009)]{kelly}
Kelly, Brandon C., Jill Bechtold, and Aneta Siemiginowska. ``Are the variations in quasar optical flux driven by thermal fluctuations?.'' The Astrophysical Journal 698, no. 1 (2009): 895.

\bibitem[Lomb(1976)]{lomb}
Lomb, Nicholas R. ``Least-squares frequency analysis of unequally spaced data.'' Astrophysics and space science 39, no. 2 (1976): 447-462.

\end{thebibliography}
